Gaussian Processes Autoregressive Models for the Disturbance Storm Time Index

### 2016-04-24 swsc.tex
Updates from ShareLaTeX

   \abstract{
    We aim to improve the state of the art of \emph{One Step Ahead}(OSA) prediction of the $D_{st}$ geomagnetic activity index. In the domain of OSA prediction, it is well known that $D_{st}(t)$ values correlate strongly with values one hour in the past (i.e. $D_{st}(t-1)$). While a number of forecasting models have been proposed and compared, none of them have been compared with the \emph{Persistence} model to quantify their performance improvements above this most trivial of predictors. In this study we train Gaussian Process regression models for One Step Ahead (OSE) prediction of the Disturbance Storm Time ($D_{st}$) geomagnetic index. We propose three variants of the Gaussian Process model.
    \begin{enumerate*}
      \item Gaussian Process Auto regressive GP-AR 
      \item Gaussian Process Auto regressive with eXogenous inputs GP-ARX. 
    \end{enumerate*}
    We compare the performance of these models with the current state of the art in one step ahead $D_{st}$ prediction on a set of 
    63 benchmark storms from 1998-2006. The proposed Gaussian Process models not only outperform the rest but also the \emph{Persistence} model.
    }

@@

@@

@@

@@

@@

\begin{figure}
   \centering
   \subfigure{\includegraphics[width=\columnwidth]{Compare_RMSE.png}}
   \vspace{12pt}
   \subfigure{\includegraphics[width=\columnwidth]{Compare_CC.png}}
   \caption{\small  Performance Comparison} 
   \label{fig:rmsecc}
   \end{figure}

\begin{figure}
   \centering
   \subfigure{\includegraphics[width=\columnwidth]{Compare_deltaDst.png}}
   \vspace{12pt}
   \subfigure{\includegraphics[width=\columnwidth]{Compare_timingerr.png}}
   \caption{\small  Performance Comparison} 
   \label{fig:deltaDst}
   \end{figure}

@@

   \begin{enumerate}
      \item Persistence model due to \emph{One Step Ahead} restriction 
      \item Superior performance of \emph{Gaussian Process} NAR and NARX models
      \item Multi step ahead prediction should be the preferred gold standard with respect to forecasting of geomagnetic indices.
   \end{enumerate}

>>>>

   \abstract{
    We aim to improve the state of the art of \emph{One Step Ahead}(OSA) prediction of the $D_{st}$ geomagnetic activity index. In the domain of OSA prediction, it is well known that $D_{st}(t)$ values correlate strongly with values one hour in the past (i.e. $D_{st}(t-1)$). While a number of forecasting models have been proposed and compared, none of them have been compared with the \emph{Persistence} model to quantify their performance improvements above this most trivial of predictors. In this study we train Gaussian Process regression models for One Step Ahead (OSE) prediction of the Disturbance Storm Time ($D_{st}$) geomagnetic index. We propose three variants of the Gaussian Process model.
    \begin{enumerate*}
      \item Gaussian Process Auto regressive (GP-AR) 
      \item Gaussian Process Auto regressive with eXogenous inputs (GP-ARX). 
    \end{enumerate*}
    We compare the performance of these models with the current state of the art in one step ahead $D_{st}$ prediction on a set of 
    63 benchmark storms from 1998-2006. The proposed Gaussian Process models not only outperform the rest but also the \emph{Persistence} model.
    }

@@

\begin{figure}[ht]
   \centering
   \subfigure{\includegraphics[width=\columnwidth]{Compare_RMSE.png}}
   \vspace{12pt}
   \subfigure{\includegraphics[width=\columnwidth]{Compare_CC.png}}
   \caption{\small  Performance Comparison; left: Averaged RMSE right: Averaged Cross correlation coefficient} 
   \label{fig:rmsecc}
   \end{figure}

@@

\begin{figure}
   \centering
   \subfigure{\includegraphics[width=\columnwidth]{Compare_deltaDst.png}}
   \vspace{12pt}
   \subfigure{\includegraphics[width=\columnwidth]{Compare_timingerr.png}}
   \caption{\small  Performance Comparison; left: Averaged $\Delta Dst_{min}$ right: Averaged absolute timing error} 
   \label{fig:deltaDst}
   \end{figure}

@@

With respect to the \emph{TL} model one important concern must be expressed. The training of the \emph{TL} model was done on Omni data from 1996-2002 which overlaps with a large number of the storm events, therefore the experimental procedure has a strong bias towards \emph{TL} model.

@@

Figures \ref{fig:rmsecc} and \ref{fig:deltaDst} compare the performance of \emph{GP-AR}, \emph{GP-ARX} and the \emph{Persistence} model with the existing results of \citet{Ji2012}. Two central themes come out of these charts.

@@

\subsection{Persistence Behavior}

It is clear that the \emph{persistence} behavior in the $D_{st}$ values is very strong i.e. the trivial predictive model $\hat{D_{st}}(t) = D_{st}(t-1)$ gives excellent performance. In fact it can be seen in these figures that the \emph{Persistence} model is outperforms every model compared by \citet{Ji2012} in the context of OSA prediction of $D_{st}$.

\subsection{Assessment of Gaussian Process Models}

The results indicate that there are many benefits to be gained by the application of Gaussian Process models for prediction of $D_{st}$. The \emph{GP-AR} and \emph{GP-ARX} not only outperform the models compared in \citet{Ji2012} but also the \emph{Persistence} model. Figure \ref{fig:relprob} compares the empirical distributions of absolute relative model errors across all storm events in the experiments. If one chooses a $95\%$ probability bound then it can be seen that the \emph{GP-ARX} model bounds the relative error to $20\%$ of the actual $D_{st}$ value, while in the case of the \emph{NM} model it bounds the relative error to $50\%$ of the actual $D_{st}$ value. Therefore the \emph{GP-ARX} model gives an approximate benefit of $25\%$ over the other compared models on the test data set chosen.

If one compares the training sets used by the models, one can appreciate that they are able to learn robust models for prediction with relatively small training and validations sets (training set contains 250 instances, while the validation set contains 432 instances) on the other hand the \emph{NM} model uses one year (8760 instances) of data to learn the formula outlined in \citet{balikhin:narmax} and the \emph{TL} model uses six years of data (52560 instances) for training.

The encouraging results of Gaussian Processes illustrate the strengths of the Bayesian approach to predictive modeling. Since the GP models generate predictive distributions for test data and not just point predictions they lend themselves to the requirements of space weather prediction very well because of the need to generate error bars on predictions.

@@

We conclude this paper with three key conclusions/insights learned from our activities so far.
   \begin{enumerate}
      \item \emph{Persistence} model must be central in the model evaluation process in the context of \emph{One Step Ahead} prediction of the $D_{st}$ index. This is due to its excellent performance in terms of errors and correlations. 
      \item \emph{Gaussian Process} AR and ARX models give encouraring benefits in OSA prediction even when compared to the \emph{Persistence} model. Leveraging the strengths of the Bayesian approach, they are able to learn robust predictors from data.
      \item Multi-step ahead prediction should be the preferred gold standard with respect to forecasting of geomagnetic indices. Also important is the usage of more informative performance metrics for evaluating and comparing empirical predictive models.
   \end{enumerate}
